---
title: "[Notebook Title Here (Goal or Example)]"
author: "[Notebook Author Name Here] ([author email here])"
date: "[Created or Edited Date Here]"
output:
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: TRUE
      smooth_scroll: false
---
```{r global options, echo = FALSE}
#List other global opts here if relevant
knitr::opts_chunk$set(fig.align="center") #Figures should be centered
```

# 1. Introduction & Overview 

[HTAN](https://humantumoratlas.org/) is a National Cancer Institute (NCI)-funded Cancer MoonshotSM initiative to construct 3-dimensional atlases of the dynamic cellular, morphological, and molecular features of human cancers as they evolve from precancerous lesions to advanced disease. [Cell April 2020](https://www.sciencedirect.com/science/article/pii/S0092867420303469) 

This notebook shows one example of how data can be accessed and analyzed using R programming software.

### 1.1 Goal
### 1.2 Inputs, Outputs, & Data (if applicable)
### 1.3 Notes (if applicable) 

# 2. Environment & Library Setup
```{r}
#List your additional libraries here
suppressMessages(library(tidyverse))
suppressMessages(library(bigrquery))
suppressMessages(library(knitr))
```

# 3. Google Authentication

Running the BigQuery cells in this notebook requires a Google Cloud Project, instructions for creating a project can be found in the [Google Documentation](https://cloud.google.com/resource-manager/docs/creating-managing-projects#console). The instance needs to be authorized to bill the project for queries. For more information on getting started in the cloud see 'Quick Start Guide to ISB-CGC' and alternative authentication methods can be found in the [Google Documentation](https://cloud.google.com/resource-manager/docs/creating-managing-projects#console).

```{r eval = FALSE}
billing <- 'your_project_number' # Insert your project ID in the ''
if (billing == 'your_project_number') {
  print('Please update the project number with your Google Cloud Project')
}

# Connect to BigQuery
con <- dbConnect(
  bigrquery::bigquery(),
  project = billing,
)
```

```{r}
#Relevant code here
```

# 4. [Main Analysis Name Here]
### 4.1 Analysis Subheader 1
```{r}
#Relevant code here
```

### 4.2 Analysis Subheader 2
```{r}
#Relevant code here
```

### 4.3 Analysis Subheader n
```{r}
#Relevant code here
```

# 5. Relevant Citations and Links (if applicable)